Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-10-17 ms_preprint/ms_preprint.tex
ms tidy-up and supplementary material in rmd

@@

\keywords{
    Assemblages
   \and
    Community ecology
   \and
    East Australian Current
   \and
    Mixture models
   \and
    Region of Common Profiles
  }

@@

The EAC originates from the South Equatorial Current (SEC), a tropical
nutrient-poor current. When the SEC's North Caledonian Jet meets the
Queensland Plateau (\textasciitilde18°S), it splits into a north-flowing
branch towards the Gulf of Papua and a south-flowing branch, which marks
the start of the EAC (Ganachaud et al., 2014; Kessler \& Cravatte,
2013). The South Caledonian Jet completes the EAC system at
\textasciitilde22°S after encountering the shelf break in the southern
portion of the Great Barrier Reef (\textbf{Fig. 1a}; Kessler \& Cravatte
(2013); Ganachaud et al. (2014)). \textbf{SST/SSS characteristics? ROB?}
The EAC interacts with the bathymetry of the seafloor, giving rise to
oceanographic features and processes, such as smaller currents,
upwellings, and persistent dynamic eddies (Oke, Roughan, et al., 2019;
Oke \& Middleton, 2000; Ridgway \& Dunn, 2003). The EAC flow polewards
and is primarily associated with the shelf break, but at
\textasciitilde32°S it detaches from the shelf and spawns an area of
eddy generation that dissipates throughout the western Tasman Sea (Mata
et al., 2006; Oke, Roughan, et al., 2019), the `EAC eastern extension'
(Oke, Pilo, et al. (2019); often termed the `Tasman Front'). Although
most EAC energy is converted into eddies, the EAC still runs southwards
following the shelf break towards Tasmania (the `EAC southern
extension'; \textbf{Fig. 1a}), carrying eddy-like features as a
transient current (Oke, Roughan, et al., 2019).

@@

Seabird data were gathered aboard \emph{RV Investigator} and \emph{RV
Falkor} on 15 voyages, between 2016/17--2020/21, spanning over
\textasciitilde37° of latitude (\textbf{Fig. 1b}; \textbf{Table S1}).
One or more observers continuously recorded seabirds within a 300 m
forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours
while the ship was underway. The species, the numbers of birds, and
their behaviours were logged in real time into a ship-based web server.
\textbf{Environmental parameters for atmospheric and oceanographic
conditions were logged continuously by automated-ship based sensors,
with seabird observations supplementing relevant environmental
conditions such as visibility or associations with marine mammals.}
Individual records received a date/time/geographic coordinate label
(multi-species records have the same label but are logged as separate
entries for each species). Birds following the vessels were excluded
from analyses. Seabird observation protocols followed Raymond et al.
(2010), in use around Australia since 1980.

To investigate and predict species' distributions and relationships with
environmental variables, a considered choice of \textbf{broad-scale
climatic and oceanographic} covariates is important (Barbet-Massin \&
Jetz, 2014). We used three physiographic variables and seven
oceanographic variables, from which a further three climatic layers were
calculated (\textbf{Table 1}). These variables are the most frequently
used in seabird-environment relationship studies (Tremblay et al.,
2009), capturing processes at a range of spatial and temporal scales
from macro-scale water-mass properties to prey availability in meso- to
submesoscale oceanographic features (Haury et al., 1978; Scales, Miller,
Hawkes, et al., 2014; Weimerskirch, 2007).

@@

Data wrangling, modelling and visualization were done in R 4.2.0 (R Core
Team, 2022); the full list of packages can be found in the Supplementary
Material \textbf{XX}. Code and the modelling dataset (see
\ref{sec:modellingdatasets}) are available in \textbf{Daudt et al.~2023
(Zenodo repo)}.

@@

The spatial and temporal persistence of oceanographic features such as
eddy kinetic energy (`eke') and thermal fronts (sea surface temperature
gradient: `sst\_grad') may influence the movements and distributions of
seabirds by offering a higher likelihood of finding food (Scales,
Miller, Embling, et al., 2014; Weimerskirch, 2007). Thus, `eke' and
`sst\_grad' environmental layers were calculated before extracting
environmental data for each seabird record. For both `eke' and
`sst\_grad', we used their original cell spatial resolution
(\textbf{Table 1}) and calculations were done layer-by-layer to match
seabird data on the finest spatial scale possible. Eddy kinetic energy
was calculated as \(0.5 (u'² + v'²)\), where \(u' = u - \bar{u}\) and
\(v' = v - \bar{v}\) are the surface zonal (latitudinal) and meridian
(longitudinal) geostrophic currents, respectively, using a rolling mean
of 90 days as \(\bar{u}\) and \(\bar{v}\) (Sun et al., 2022). The
spatial `sst\_grad' for each cell was calculated by averaging the
maximum value for each neighbouring cell (average neighbourhood
maximum), from north-to-south (NS) and west-to-east (WE); then, the
magnitude of the gradient was defined as \(NS² + WE²\) (Burrows et al.,
2011). In addition, we calculated climatic seasonal average layers for
`eke' and `sst\_grad' based on the 11-year data for 2011--2021 as a
proxy for persistent oceanographic features, and the `eke' standard
deviation over the same period (named with the prefix `clim\_' in
\textbf{Table 1}). Seabed slope was calculated using `raster::terrain()'
(Hijmans, 2022a), and distance from the coast was calculated through
`sf::st\_distance()' (Pebesma, 2018) using Australia and Tasmania as the
mainland (i.e.~ignoring smaller islands). Then, for each seabird record,
the most proximate value for each environmental layer (original or
calculated) was extracted using built-in functions from `raster' 3.5-21,
`terra' 1.6-7, and `sf' 1.0-8 packages (Hijmans, 2022a, 2022b; Pebesma,
2018).

@@

Seabirds are highly mobile animals, and many species migrate and/or
disperse during their non-breeding periods (Schreiber \& Burger, 2002).
Therefore, seasonal differences in species composition were expected
(e.g. K. Reid et al., 2023), and in the probabilities of occurrence and
in total numbers. Thus, we built separate modelling dataset for each
season. We aggregated seabird data by 1° latitude \(\times\) 1°
longitude grid cells (c.~110 \(\times\) 110 km) and season (\textbf{Fig.
1b}), by summing seabirds from the same species. To test if the
presence-absence and abundance (raw counts) data would result in similar
assemblage patterns, we built two species matrices, where the
presence-absence matrix transformed values \textgreater1 to 1. To avoid
numerical instabilities in the models prompted by unusual or rare
species, we kept only species occurring in at least six grids within
each season (summer = 12 species; autumn = 21; winter = 20; spring =
33). The associated environmental data of each seabird record were
averaged within the same grid cells and seasons, and environmental data
were standardised before modelling.

@@

A total of 10,261 seabird records was obtained during the 15 voyages,
comprising 142,646 individuals from 80 species, belonging to 11 families
and five orders. The five most abundant species made up 83\% of total
birds counted (43.0\% short-tailed shearwater \emph{Ardenna
tenuirostris}, 17.2\% fairy prion \emph{Pachyptila turtur}, 14.3\%
wedge-tailed shearwater \emph{Ardenna pacifica}, 5.2\% sooty tern
\emph{Onychoprion fuscatus}, 3.2\% shy albatross \emph{Thalassarche
cauta}). The five most frequently encountered species were \emph{A.
pacifica} (frequency of occurrence = 35.3\%), \emph{T. cauta} (33.2\%),
brown booby \emph{Sula leucogaster} (25.4\%), red-footed booby \emph{S.
sula} (25.4\%) and black-browed albatross \emph{T. melanophris} (25\%)
(\textbf{Fig. S1}). An increase in species richness towards the south of
the study region was clear in all but the winter season (\textbf{Fig
S2}). Noticeably higher counts were made around Tasmania, possibly due
to the `colony effect' and/or the presence of Southern Ocean species
(\textbf{Fig. S2}). However, one cell in the Coral Sea had counts of the
same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups,
except for the presence-absence-autumn model, which indicated three RCPs
as the best fit (\textbf{Fig. S3}). The best fitting models generally
had 1 or 2 covariates selected, although the abundance-autumn model had
five covariates (\textbf{Table 2}). The best models showed a good fit
with no apparent deviations from the assumptions (\textbf{Fig. S4}). All
the best models selected `sst' or `sss', as essential for characterising
water masses. The probability of belonging to one or another RCP varied
its `sst' threshold between 18--21 °C, or `sss' 35.1 PSU for summer
(\textbf{Fig. 2}, \textbf{Fig. S5}), similar to the EAC's temperature
and salinity thresholds. Furthermore, most of the best models selected
`bat', suggesting a gradient between `coastal' and `oceanic' taxa based
on bathymetry (\textbf{Fig. S5}). Of particular note, proxies for
persistent oceanographic features that may aggregate food resources (the
`climatic' variables) for seabirds were selected only in the autumn and
spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest
probability value for a given cell. When predicting RCP onto space,
patterns of point-predictions suggest a biogeographic transition in
seabird species around \textasciitilde34°S, with both presence-absence
and abundance models pointing to similar results (\textbf{Fig. 3}).
These patterns are further reinforced when looking at the mean
probability along with the lower and upper confidence intervals of
predictions (\textbf{Fig. 4}, \textbf{Fig. S6}), showing that around
\textasciitilde34°S, the `northern' RCP tended to have higher
probabilities in coastal regions, whereas the `southern' RCP spread
northward offshore. The presence-absence-autumn model (three RCPs) had a
similar broad-scale pattern, although the southern RCP is farther south,
and a strong `Tasmania/Bass Straight' assemblage appears (\textbf{Fig.
3}, \textbf{Fig. S6}).

The broad pattern of two large-scale assemblages is reflected in their
species compositions. Although there are overlaps among the seabird
groups, both presence-absence and abundance models captured typical
tropical species in the northern RCP (e.g.~boobies \emph{Sula} spp.,
lesser frigatebird \emph{Fregata ariel}) and typical Southern Ocean
species in the southern RCP (e.g.~small albatrosses \emph{Thalassarche}
spp., fairy prion \emph{P. turtur}) (\textbf{Fig. 5}, \textbf{Fig. S7}).
The presence-absence-autumn model, which had three RCPs, also showed the
characteristic taxa representing the `northern' and `southern' RCPs.
However, the intermediate `Tasmania/Bass Straight' assemblage showed a
mixture of both groups with shy albatross and Australasian gannet
\emph{Morus serrator} prominent with high probabilities of occurrence.

>>>>

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

\keywords{
    Assemblages
   \and
    Biogeography
   \and
    Community ecology
   \and
    East Australian Current
   \and
    Mixture models
   \and
    Region of Common Profiles
  }

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\textasciitilde18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC (Ganachaud et al., 2014; Kessler \& Cravatte, 2013). The South Caledonian Jet completes the EAC system at \textasciitilde22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (\textbf{Fig. \ref{fig:fig1-study-area}a}; Kessler \& Cravatte (2013); Ganachaud et al. (2014)). \textbf{SST/SSS characteristics? ROB?} The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies (Oke, Roughan, et al., 2019; Oke \& Middleton, 2000; Ridgway \& Dunn, 2003). The EAC flow polewards and is primarily associated with the shelf break, but at \textasciitilde32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea (Mata et al., 2006; Oke, Roughan, et al., 2019), the `EAC eastern extension' (Oke, Pilo, et al. (2019); often termed the `Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the `EAC southern extension'; \textbf{Fig. \ref{fig:fig1-study-area}a}), carrying eddy-like features as a transient current (Oke, Roughan, et al., 2019).

@@

Seabird data were gathered aboard \emph{RV Investigator} and \emph{RV Falkor} on 15 voyages, between 2016/17--2020/21, spanning over \textasciitilde37° of latitude (\textbf{Fig. \ref{fig:fig1-study-area}b}; \textbf{Table S1}). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed Raymond et al. (2010), in use around Australia since 1980.

To investigate and predict species' distributions and relationships with environmental variables, a considered choice of broad-scale climatic and oceanographic covariates is important (Barbet-Massin \& Jetz, 2014). We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (\textbf{Table \ref{tab:table1}}). These variables are the most frequently used in seabird-environment relationship studies (Tremblay et al., 2009), capturing processes at a range of spatial and temporal scales from macro-scale water-mass properties to prey availability in meso- to submesoscale oceanographic features (Haury et al., 1978; Scales, Miller, Hawkes, et al., 2014; Weimerskirch, 2007).

@@

Data wrangling, modelling and visualization were done in R 4.2.0 (R Core Team, 2022); the full list of packages can be found in the Supplementary Material. Code and the modelling dataset (see \ref{sec:modellingdatasets}) are available in \textbf{(Zenodo repo)}.

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy (`eke') and thermal fronts (sea surface temperature gradient: `sst\_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food (Scales, Miller, Embling, et al., 2014; Weimerskirch, 2007). Thus, `eke' and `sst\_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both `eke' and `sst\_grad', we used their original cell spatial resolution (\textbf{Table \ref{tab:table1}}) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as \(0.5 (u'² + v'²)\), where \(u' = u - \bar{u}\) and \(v' = v - \bar{v}\) are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as \(\bar{u}\) and \(\bar{v}\) (Sun et al., 2022). The spatial `sst\_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as \(NS² + WE²\) (Burrows et al., 2011). In addition, we calculated climatic seasonal average layers for `eke' and `sst\_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the `eke' standard deviation over the same period (named with the prefix `clim\_' in \textbf{Table \ref{tab:table1}}). Seabed slope was calculated using `raster::terrain()' (Hijmans, 2022a), and distance from the coast was calculated through `sf::st\_distance()' (Pebesma, 2018) using Australia and Tasmania as the mainland (i.e.~ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from `raster' 3.5-21, `terra' 1.6-7, and `sf' 1.0-8 packages (Hijmans, 2022a, 2022b; Pebesma, 2018).

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods (Schreiber \& Burger, 2002). Therefore, seasonal differences in species composition were expected (e.g. K. Reid et al., 2023), and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude \(\times\) 1° longitude grid cells (c.~110 \(\times\) 110 km) and season (\textbf{Fig. \ref{fig:fig1-study-area}b}), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \textgreater1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83\% of total birds counted (43.0\% short-tailed shearwater \emph{Ardenna tenuirostris}, 17.2\% fairy prion \emph{Pachyptila turtur}, 14.3\% wedge-tailed shearwater \emph{Ardenna pacifica}, 5.2\% sooty tern \emph{Onychoprion fuscatus}, 3.2\% shy albatross \emph{Thalassarche cauta}). The five most frequently encountered species were \emph{A. pacifica} (frequency of occurrence = 35.3\%), \emph{T. cauta} (33.2\%), brown booby \emph{Sula leucogaster} (25.4\%), red-footed booby \emph{S. sula} (25.4\%) and black-browed albatross \emph{T. melanophris} (25\%) (\textbf{Fig. S1}). An increase in species richness towards the south of the study region was clear in all but the winter season (\textbf{Fig. S2}). Noticeably higher counts were made around Tasmania, possibly due to the `colony effect' and/or the presence of Southern Ocean species (\textbf{Fig. S2}). However, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania in summer.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (\textbf{Fig. S3}). The best fitting models generally had 1 or 2 covariates selected, although the abundance-autumn model had five covariates (\textbf{Table \ref{tab:table2}}). The best models showed a good fit with no apparent deviations from the assumptions (\textbf{Fig. S4}). All the best models selected `sst' or `sss', as essential for characterising water masses. The probability of belonging to one or another RCP varied its `sst' threshold between 18--21 °C, or `sss' 35.1 PSU for summer (\textbf{Fig. \ref{fig:fig2-partial-plots}}, \textbf{Fig. S5}), similar to the EAC's temperature and salinity thresholds. Furthermore, most of the best models selected `bat', suggesting a gradient between `coastal' and `oceanic' taxa based on bathymetry (\textbf{Fig. S5}). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the `climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \textasciitilde34°S, with both presence-absence and abundance models pointing to similar results (\textbf{Fig. \ref{fig:fig3-point-pred}}). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (\textbf{Fig. \ref{fig:fig4-prob-maps}}, \textbf{Fig. S6}), showing that around \textasciitilde34°S, the `northern' RCP tended to have higher probabilities in coastal regions, whereas the `southern' RCP spread northward offshore. The presence-absence-autumn model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong `Tasmania/Bass Straight' assemblage appears (\textbf{Fig. \ref{fig:fig3-point-pred}}, \textbf{Fig. S6}).

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g.~boobies \emph{Sula} spp., lesser frigatebird \emph{Fregata ariel}) and typical Southern Ocean species in the southern RCP (e.g.~small albatrosses \emph{Thalassarche} spp., fairy prion \emph{P. turtur}) (\textbf{Fig. \ref{fig:fig5-spp-profiles}}, \textbf{Fig. S7}). The presence-absence-autumn model, which had three RCPs, also showed the characteristic taxa representing the `northern' and `southern' RCPs. However, the intermediate `Tasmania/Bass Straight' assemblage showed a mixture of both groups with shy albatross and Australasian gannet \emph{Morus serrator} prominent with high probabilities of occurrence.
